---
title: 'W203 Lab 1: Exploratory Data Analysis of Corruption and Parking Violations'
author: "Chi Iong Ansjory, Tsung-Chin Han, Marcelo Queiroz"
date: 5/25/2018
output: pdf_document
---



#Overview

The goal of this lab is to provide you with an opportunity to use R and gain experience performing exploratory data analysis (EDA). In this lab, you will be asked to find new insight into a data set by assessing the underlying structure, evaluating the variables, detecting outliers and anomalies, and so on. 

This is a group lab.  Each team has been assigned a different dataset to work on.  In your assigned folder, you will find a file containing background information on your data, along with a research objective and any instructions that are specific to your team.

Please note that you will be working with real data, but it may have been modified by your instructors to test your abilities.

Although your assigned topic may be the focus of an existing literature, we recommend that you do not spend your time researching what others have done, or gaining significant domain expertise.  The purpose of the lab is to see how well you can apply exploratory techniques.  Moreover, the background we have provided in your assignment should be sufficient to guide your analysis.\footnote{We also do not want you to be led astray by the bad advice that is common on the internet.}

#Assignment

Generate an exploratory analysis to address the goals found in your assigned folder.

Be sure to follow the guidelines we covered in class.  Remember that you are to use descriptive tools (no inference), but note any features you find that you think would be relevant to statistical modeling.

Your analysis should be thorough, but limit your report to a maximum of 25 pages.  This means that you will have to make choices about what variables and relationships to focus on (and justify those choices).

To assist with evaluation, we are providing the following outline for your report.  As you work, you may fill in each section with your analysis.

### Introduction (20 pts)

#### State the research question that motivates your analysis.

How is the amount of parking violations received by a UN diplomat related to the country corruption index?

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir='~/Documents/UC Berkeley MIDS/Summer 2018/DATASCI W203/Lab_1/Corruption_EDA')
```

#### Load your data set into R. 
```{r}
load("Corrupt.RData")
```

#### Describe your data set.  What types of variables does it contain?  How many observations are there?

We note that we have 364 observations and 28 variables.
```{r}
nrow(FMcorrupt)
```

```{r}
str(FMcorrupt)
```

#### Evaluate the data quality. Are there any issues with the data? Explain how you handled these potential issues. 
```{r}
summary(FMcorrupt$violations)
```

```{r}
summary(FMcorrupt$corruption)
```

#### Explain whether any data processing or preparation is required for your data set. 

```{r}
subcase_nona = !is.na(FMcorrupt$corruption)
FMcorrupt_nona = FMcorrupt[subcase_nona, ]
nrow(FMcorrupt_nona)
```

### Univariate Analysis of Key Variables (20 pts)

#### Use visualizations and descriptive statistics to perform a univariate analysis of each key variable. Be sure to describe any anomalies, coding issues, or potentially erroneous values.  Explain how you respond to each issue you identify.  Note any features that appear relevant to statistical analysis.  Discuss what transformations may be appropriate for each variable.
```{r car}
library(car)
```

```{r}
scatterplotMatrix(~ violations + corruption + trade + region, 
                  data = FMcorrupt_nona, 
                  main = "Scatterplot Matrix for Key Variables")
```

### Analysis of Key Relationships (30 pts)

#### Explore how your outcome variable is related to the other variables in your dataset. Make sure to use visualizations to understand the nature of each bivariate relationship.

First, use plot to figure out the relationship between violations and corruptions. From the plot, the majority of density of violations take place in the corruptions between 0 and 1.
```{r}
plot(jitter(FMcorrupt_nona$corruption, factor=2), jitter(FMcorrupt_nona$violations, factor=2), xlab = "corruptions", ylab = "violations")
abline(lm(FMcorrupt_nona$violations ~ FMcorrupt_nona$corruption))
```

Then, inspect the relationship between violations and majoritymuslim. There is no clear correlation between them.
```{r}
plot((FMcorrupt_nona$majoritymuslim),(FMcorrupt_nona$violations), xlab = "majoritymuslim", ylab = "violations")
abline(lm(FMcorrupt_nona$violations ~ FMcorrupt_nona$majoritymuslim))
```

Now, trying to plot the corruptions against each region. Region 6 has the most distribution of corruptions between 0 and 1.
```{r}
plot((FMcorrupt_nona$region), (FMcorrupt_nona$corruption), xlab = "region", ylab = "corruptions")
abline(lm(FMcorrupt_nona$corruption ~ FMcorrupt_nona$region))
```

Region 6 tends to have lower total trade with United States. 
```{r}
plot(FMcorrupt_nona$region, log(FMcorrupt_nona$trade), xlab = "region", ylab = "trade")
abline(lm(FMcorrupt_nona$trade ~ FMcorrupt_nona$region))
```

Region 6 has the most UN diplomats presence.
```{r}
hist(FMcorrupt_nona$region, breaks = 0:8 - 0.5, xlab = "region")
axis(1, at = 0:8)
```

### What tranformations can you apply to clarify the relationships you see in the data?  Be sure to justify each transformation you use.

Now, use transformation to find relationship between region number and the region name:

Region 6 = Africa
```{r}
subcase_africa = FMcorrupt_nona$r_africa == 1 & !is.na(FMcorrupt_nona$r_africa)
FMcorrupt_africa = FMcorrupt_nona[subcase_africa, ]
summary(FMcorrupt_africa$region)
```

Region 7 = Middle East
```{r}
subcase_middleeast = FMcorrupt_nona$r_middleeast == 1 & !is.na(FMcorrupt_nona$r_middleeast)
FMcorrupt_middleeast = FMcorrupt_nona[subcase_middleeast, ]
summary(FMcorrupt_middleeast$region)
```

Region 3 = Europe
```{r}
subcase_europe = FMcorrupt_nona$r_europe == 1 & !is.na(FMcorrupt_nona$r_europe)
FMcorrupt_europe = FMcorrupt_nona[subcase_europe, ]
summary(FMcorrupt_europe$region)
```

Region 2 = South America
```{r}
subcase_southamerica = FMcorrupt_nona$r_southamerica == 1 & !is.na(FMcorrupt_nona$r_southamerica)
FMcorrupt_southamerica = FMcorrupt_nona[subcase_southamerica, ]
summary(FMcorrupt_southamerica$region)
```

Region 4 = Asia
```{r}
subcase_asia = FMcorrupt_nona$r_asia == 1 & !is.na(FMcorrupt_nona$r_asia)
FMcorrupt_asia = FMcorrupt_nona[subcase_asia, ]
summary(FMcorrupt_asia$region)
```

### Analysis of Secondary Effects (10 pts)

#### What secondary variables might have confounding effects on the relationships you have identified?  Explain how these variables affect your understanding of the data.

The variable prepost can further segment the data set to see if there is any effect of legal penalities against the violations after 2002.
```{r}
subcase_pre = FMcorrupt_nona$prepost == "pre"
FMcorrupt_pre = FMcorrupt_nona[subcase_pre, ]
nrow(FMcorrupt_pre)
```

```{r}
summary(FMcorrupt_pre)
```

```{r}
subcase_post = FMcorrupt_nona$prepost == "pos"
FMcorrupt_post = FMcorrupt_nona[subcase_post, ]
nrow(FMcorrupt_post)
```

```{r}
summary(FMcorrupt_post)
```

The overall trends between Pre 2002 and Post 2002 are basically the same. However, the overall violations are comparatively lower Post 2002 with same trend as Pre 2002.
```{r}
scatterplotMatrix(~ violations + corruption + trade + region, 
                  data = FMcorrupt_pre, 
                  main = "Scatterplot Matrix for Key Variables Pre 2002")
```

```{r}
scatterplotMatrix(~ violations + corruption + trade + region, 
                  data = FMcorrupt_post, 
                  main = "Scatterplot Matrix for Key Variables Post 2002")
```

### Conclusion (20 pts)
#### Summarize your exploratory analysis. What can you conclude based on your analysis?  

The major contribution to diplomatic behavior is cultural norm and legal penalties don't alter the behavior but only suppress the violations. The diplomats from Africa have the significant contribtions to the violations indicated by the Country corruption index between 0 and 1, and lower total trade with the United States. The amount of African diplomats is the most among diplomats from other regions.
```{r}

```

# Evaluation

We will evaluate your report for technical correctness, but also clarity and overall effectiveness.  A point distribution is provided with the above outline.  In addition to these point totals, we will impose penalties for output dumps, unclear language, and other errors.


# Submission

Only one student in the team needs to submit via the ISVC.  Make sure that you include the names of all group members in your report.

You must turn in
\begin{enumerate}
\item Your pdf report.  In this report, do not suppress the R code that generates your output.
\item The source file you use to generate your report (i.e. your Rmd file)
\end{enumerate}

Use the following naming convention for your files:

- lastname1_lastname2_lab1.pdf
- lastname1_lastname2_lab1.Rmd


## Due Date

This lab is due 24 hours before the week 4 live session.

## Presentation

During your week 4 live session, your team should present your analysis to the class.  Please limit your presentation to 15 minutes (10 minutes plus 5 minutes for questions).  You should use this presentation to highlight the process you followed in your EDA, as well as any aspects of your data that find particularly interesting.

